---
title: Random Forest for Trees
author: Willian Vieira
date: "`r paste('Last updated on', format(Sys.time(), '%d %B, %Y'))`"

---

```{r load R packages, echo=FALSE}
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(tidyr))
suppressPackageStartupMessages(library(dplyr))
```

```{r getting variables, echo=FALSE}

  # species_id
  sp_ids = as.character(read.table('data/sp_ids.txt')[, 1])

  # other variables from the simulation file
  rFile <- readLines('R/03_runSimulations.R')

  # variables
  eval(parse(text =
    rFile[which(sapply(rFile, function(x) grep('variables = ', x)) == 1)]))

  # nbTrees
  eval(parse(text =
    rFile[which(sapply(rFile, function(x) grep('nbTrees = ', x)) == 1)]))

  # nbMtry
  eval(parse(text =
    rFile[which(sapply(rFile, function(x) grep('nbMtry = ', x)) == 1)]))

```

# Introduction

Here I use the random forest method to extract the power of different variables to explain **growth** and **mortality** of tree species.
I use the forest inventory data base for the eastern North American forest that contains... #TODO

```{r explanatory variables, echo=FALSE}
  vars = as.character(read.table('data/variables1.txt')[, 1])
```

For each tree species in the data base (total of `r length(sp_ids)`), we will run a random forest for growth and mortality with different parameters and explanatory variables.
Then we evaluate the simulation based on the R$^2$ for the growth and ... #TODO
We further check which variables have a better explanatory power over the response variable.

# Simulations

For each tree species, we ran a random forest varying (i) the explanatory variables, and (ii) the number of variables to possibly split at in each node (`mtry`).
The `r length(variables)` sets of variables are the following:

```{r table of variables, echo=FALSE}
# get variables
for(var in variables) assign(paste0('var', var), read.table(paste0('data/variables', var, '.txt'))[, 1])
for(var in variables) {print(paste('set', var)); print(as.character(get(paste0('var', var)))); cat('\n')}
```

For each set of variables, `mtry` varied as `r nbMtry`, with a fixed number of `r nbTrees` trees.

# Summary

```{r load simulations, echo=FALSE}
for(vital in c('growth', 'mort')) {
  for(sp in sp_ids) {
    for(var in variables) {
      for(tree in nbTrees) {
        for(mty in nbMtry) {
          simName = paste0(vital, sp, var, tree, mty)
          simLoc = paste0('output/', vital, '_', sp, '_var', var, '_nTrees', nbTrees, '_Mtry', mty, '.RDS')
          assign(simName, readRDS(simLoc))
        }
      }
    }
  }
}
```

```{r prepare data, echo=FALSE}

df <- expand.grid(c('growth', 'mort'), sp_ids, variables, nbTrees, nbMtry)
names(df) <- c('vital', 'sp', 'var', 'nbTrees', 'nbMtry')
df$var <- as.factor(df$var)
df$nbTrees <- as.factor(df$nbTrees)
df$nbMtry <- as.factor(df$nbMtry)

# get explanatory power
getPower <- function(x)
{
  sim = get(paste(x, sep = '', collapse = ''))
  simType = sim$treetype

  if(simType == 'Regression') {
    pw = sim$r.squared
  }else if(simType == 'Classification') {
    pw = 1 - sim$prediction.error
  }
  return(pw)
}

df$pw <- apply(df, 1, getPower)

# get importance
import <- as.data.frame(matrix(rep(NA, length(var1) * nrow(df)), ncol = length(var1)))
names(import) <- var1
df <- cbind(df, import)

simNames <- apply(df[, 1:5], 1, paste, collapse = '')
for(i in 1:nrow(df)) {
  impRF <- ranger::importance(get(simNames[i]))
  df[i, names(impRF)] <- impRF
}

# tranform to long format
dfLong = gather(df, variable, importance, growth:tot_pp_period3_lag, factor_key = TRUE)
```


## Variables set

```{r variables set, echo=FALSE}
p = ggplot(data = df) +
    aes(var, pw) +
    geom_boxplot() +
    facet_grid(~vital,
              labeller = labeller(vital = setNames(c('growth (Rsquared)',
                                  'mort (1 - OOB prediction error)'),
                                  c('growth', 'mort')))) +
    theme_classic() +
    xlab("Variable set") +
    ylab("Explanatory power of variables")

print(p)
```

Assuming the set of variables `var2` was the best to explain growth and mortality, let's see the importance of the variables present in the set `var2`:

```{r variable set importance, echo=FALSE}
p = ggplot(data = droplevels(subset(dfLong, var == 2 & variable %in% var2))) +
    aes(reorder(variable, importance, FUN = median), importance) +
    geom_boxplot() +
    coord_flip() +
    facet_grid(~vital, scales = "free_x") +
    theme_classic() +
    xlab("Variables in set 2") +
    ylab("Importance")

suppressWarnings(print(p))
```

Now for the other two sets of variables (`var1` and `var3`):

```{r variable set importance2, echo=FALSE}
p = ggplot(data = droplevels(subset(dfLong, var == 1 & variable %in% var1))) +
    aes(reorder(variable, importance, FUN = median), importance) +
    geom_boxplot() +
    coord_flip() +
    facet_grid(~vital, scales = "free_x") +
    theme_classic() +
    xlab("Variables in set 1") +
    ylab("Importance")

suppressWarnings(print(p))
```

```{r variable set importance3, echo=FALSE}
p = ggplot(data = droplevels(subset(dfLong, var == 3 & variable %in% var3))) +
    aes(reorder(variable, importance, FUN = median), importance) +
    geom_boxplot() +
    coord_flip() +
    facet_grid(~vital, scales = "free_x") +
    theme_classic() +
    xlab("Variables in set 3") +
    ylab("Importance")

suppressWarnings(print(p))
```

## Number of variables to possibly split at in each node

```{r mtry, echo=FALSE}
p = ggplot(data = df) +
    aes(nbMtry, pw) +
    geom_boxplot() +
    facet_grid(~vital,
              labeller = labeller(vital = setNames(c('growth (Rsquared)',
                                  'mort (1 - OOB prediction error)'),
                                  c('growth', 'mort')))) +
    theme_classic() +
    xlab("Number of variables to possibly split at in each node (mtry)") +
    ylab("Explanatory power of variables")

print(p)
```

## Individual species response

```{r species_id, echo=FALSE}
p = ggplot(data = df) +
    aes(reorder(sp, pw, FUN = median), pw) +
    geom_boxplot() +
    coord_flip() +
    facet_grid(~vital,
              labeller = labeller(vital = setNames(c('growth (Rsquared)',
                                  'mort (1 - OOB prediction error)'),
                                  c('growth', 'mort')))) +
    theme_classic() +
    xlab("Species id") +
    ylab("Explanatory power of variables")

print(p)
```


```{r species_id vs n, echo=FALSE}

pwSp <- df %>%
        group_by(vital, sp) %>%
        summarise(median = median(pw))

getSampleSize <- function(x)
{
  sim = simNames[grep(paste(x, sep = '', collapse = ''), simNames)[1]]
  return (get(sim)$num.samples)
}

pwSp$sampleSize <- apply(pwSp[, 1:2], 1, getSampleSize)


p = ggplot(data = pwSp) +
    aes(sampleSize, median) +
    geom_point() +
    geom_smooth(method='lm', formula= y ~ x + I(x^2)) +
    facet_grid(~vital,
              labeller = labeller(vital = setNames(c('growth (Rsquared)',
                                  'mort (1 - OOB prediction error)'),
                                  c('growth', 'mort')))) +
    theme_classic() +
    xlab("Sample size") +
    ylab("Explanatory power of variables")

print(p)
```
